A Mountaintop View Requires Minimal Sorting: \break A Faster Contour Tree Algorithm \footnote{The full updated version of this paper is available online \url{\si{http://web.engr.illinois.edu/\string~\si{raichel2}/\si{paint.pdf}}}}

### 2014-11-04 paint-contour-tree.tex
more small changes

We present the first algorithm that can avoid the global sort and has a refined time complexity that
depends on the contour tree structure. Intuitively, if the tree is short and fat, we get significant improvements 
in running time. For a partition of the contour tree into a set of descending paths, $P$,
our algorithm runs in $O(\sum_{p\in P} |p|\log|p| + t\alpha(t)+N)$, which is 
always bounded by $O(t\log t + N)$ and is $O(t\alpha(t)+N)$ for balanced trees. 
Moreover, we prove a strong optimality property of our algorithm, showing that $\sum_{p\in P} |p|\log|p|$ complexity
is inherent to computing contour trees. Consider the family of instances where our algorithm
requires $C$ comparison. Any algorithm computing contour trees on this family requires
$\Omega(C)$ comparisons in the worst case.

@@

\paragraph{Organization.}  
\Sec{basics} contains basic definition for contour trees.  In \Sec{approach} we give a high level overview and intuition for our approach 
for computing contour trees.  \Sec{surgery} and \Sec{rain} describe how we break the intial input into nicely behaved (extemum dominant) pieces. 
\Sec{extreme} shows that computing the contour tree for these nicely behaved pieces is roughly equivalent to computing the join tree. 
\Sec{paint} contains our algorithm for computing join trees, which along with previous sections gives our algorithm for contour trees.  
Finally, in \Sec{pathDecomp} and \Sec{lb} we relate the running time of our algorithm (providing upper and lower bounds) to a particular decomposition of the contour tree.

@@

The $(f(x)+\delta)$-contours in $f|_{B_\eps(x)}$ given by \Def{deg} might actually be the same
contour in $f$. The up-degree (as opposed to \emph{Morse} up-degree) is defined as the number of $(f(x)+\delta)$-contours
that intersect $B_\eps(x)$, a potentially smaller number. This up-degree
is exactly the up-degree of $x$ in $\reeb(\MM)$. (Analogously, for down-degree.)
When the Morse up-degree is $2$ but the up-degree is $1$, the topology of the level set changes but not by the number of connected components changing.
For example, when $d=3$ this is equivalent to the contour gaining a handle.
When $d=2$, this distinction is not necessary, since any point with $>1$
Morse degree will have some $>1$ degree in $\reeb(\MM)$.

@@

\begin{figure}[h]\centering
    \includegraphics[width=.18\linewidth]{figs/tent2}%
    \hspace{.9in}
    \includegraphics[width=.18\linewidth]{figs/tent1}
    \caption{On left, downward rain spilling only (each shade of gray represents a different color), producing a grid.  
    Note we are assuming raining was done in reverse sorted order of maxima.  
    On right, flipping the direction of rain spilling.}
    \label{fig:order}
\end{figure}

@@

We stress the importance of this step for our approach. Given the algorithm of~\cite{csa-cctad-00},
one may think that it suffices to design faster algorithms for join trees. But this
cannot give the sort of optimality we hope for. Again, consider \Fig{sorting}. Any algorithm
to construct the true join tree must construct the path of splits, which implies
sorting all of them. It is absolutely necessary to cut $\MM$ into pieces where the cost
of building the join tree can be related to the cost of building $\cC(\MM)$.

@@

\paragraph{The running time analysis:} Relating the running time to a path decomposition is the most technical part of the paper. 
All the non-heap operations can be easily bounded as $O(t\alpha(t) + N)$ ($t\alpha(t)$ from the union-find data structure
for colors). It is not hard to argue that at all times, any heap always contains a subset of a leaf
to root path. Unfortunately, this subset is \emph{not} contiguous, as \Fig{colors} shows. \Sesh{explain}
Nonetheless, we can get a non-trivial (but non-optimal) bound. Let $d_v$ denote the distance to the root for vertex $v$ in the join tree.
The total cost (of the heap operations) is at most $\sum_v \log d_v$. This immediately proves a bound of $O(t\log D)$, where
$D$ is the maximum distance to the root, an improvement over previous work. But this bound is non-optimal.
For a balanced binary tree, this bound is $O(t\log\log t)$, whereas the cost of any path decomposition is $O(t)$.

@@

Consider a contour tree $T$ and the path decomposition $P(T)$ used to bound the running time. 
Denoting $\cost(P(T)) = \sum_{p \in P(T)} |p|\log |p|$, we construct a set of $\prod_{p \in P(T)} |p|!$
functions on a fixed domain such that each function has a distinct (labeled) contour tree. By a simple
entropy argument, any height decision tree that correctly computes the contour tree on all instances
requires average case $\Omega(\cost(P(T)))$ time. We prove that our algorithm makes $\Theta(\cost(P(T)))$
height comparisons on all these instances. In general, any algorithm requires $\Omega(C)$ time
to solve the set of instances that our algorithm solves with $C$ comparisons.

@@

\section{Acknowledgements}

We thank Jeff Erickson and Yusu Wang for numerous useful discussions. This work is
supported by the Laboratory Directed Research and Development
(LDRD) program of Sandia National Laboratories. Sandia National
Laboratories is a multi-program laboratory managed and operated
by Sandia Corporation, a wholly owned subsidiary of Lockheed
Martin Corporation, for the U.S. Department of Energy’s National
Nuclear Security Administration under contract DE-AC04-94AL85000.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
%end of writup
%stuff below should be deleted
%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\remove{

@@

\section{Old stuff}

\subsubsection{Paying it down the line}

% \begin{claim}
% \label{clm:sizes}
%  Let $p$ be a path in a maximum path decomposition, such that $|H_p|>c|p|$ for some constant $c> 6$.
%  Let $H_f$ denote the heap at the leaf of $p$.  We then have the following
%  \begin{compactenum}
%   \item $|H_f|\leq (1+1/c)|H_p|$
%   \item $|H_f|\geq |H_p|/3$
%   \item For any path $q$ adjacent to $p$ in the shrub, $|H_q|<\frac{2}{3}|H_p|$
%  \end{compactenum} 
% \end{claim}
% \begin{proof}
%   For any vertex $v\in p$ \Obs{decrease} implies that $|H_v|\leq |H_p| + |p| < (1+1/c) |H_p|$.
%   First observe that by setting $v=f$ this immediately implies the first part of the claim.
%  % By a similar argument, $|H_f|\geq |H_p| - |p| > (1-1/c)|H_p|$.
%   
%   To prove the last part of the claim, suppose otherwise that $|H_q|\geq \frac{2}{3}|H_p|$ for some 
%   adjacent path $q$.  As this is a maximum path decomposition, this implies that there is some 
%   $v\in p$ such that $|H_v|\geq \frac{4}{3}|H_p|-1$.
%   By \Obs{decrease} we then have, 
%    $|H_p| \geq |H_v| - |p|\geq \frac{4}{3}|H_p| - 1 - |H_p|/c \geq (4/3-1/c)|H_p|-1$.
%   When rearranged this gives $1\geq (1/3-1/c)|H_p|$.  As $c>6$ we have $1\geq |H_p|/6$, 
%   which is a contradiction as $|H_p|>c|p|>6$.
% \end{proof}

\Ben{I had originally thought the following claim would be useful in proving the lemma below.}
\begin{claim}
\label{clm:sizes}
 Let $p$ be a path in a maximum path decomposition, such that $|H_p|>c|p|$ for some constant $c> 6$.
 We then have the following
% \begin{compactenum}
%  \item 
  For any path $q$ adjacent to $p$ in the shrub, $|H_q|<\frac{2}{3}|H_p|$
%  \item If there does not exist and adjacent path $q$ such $H_q\geq H_p/3$, then $|H_f|\geq |H_p|/3$
% \end{compactenum} 
\end{claim}
\begin{proof}
  To prove the claim, suppose otherwise that $|H_q|\geq \frac{2}{3}|H_p|$ for some 
  adjacent path $q$.  As this is a maximum path decomposition, this implies that there is some 
  $v\in p$ such that $|H_v|\geq \frac{4}{3}|H_p|-1$.
  By \Obs{decrease} we then have, 
   $|H_p| \geq |H_v| - |p|\geq \frac{4}{3}|H_p| - 1 - |H_p|/c \geq (4/3-1/c)|H_p|-1$.
  When rearranged this gives $1\geq (1/3-1/c)|H_p|$.  As $c>6$ we have $1\geq |H_p|/6$, 
  which is a contradiction as $|H_p|>c|p|>6$.  
\end{proof}

\begin{lemma}
\label{lem:excess}
 Let $p$ be a tall path, and let $S_1, \dots S_k$ and $R_1, \dots, R_l$ be the sizes of the heaps at the roots of the short and tall children of $p$, respectively, in the shrub.  
 Let $H_f$ be the heap at the leaf of the path $p$.
 Then,  $|H_p| \log |H_p|   \geq  \sum_{v\in p} \log|H_v| + \frac{4}{3}\sum_{i=1}^k S_i \log S_i + \sum_{i=1}^l R_i \log R_i  -  c|p|\log |p|$, for some sufficiently large constant $c$
 \end{lemma}
\begin{proof}
 First observe that the lemma holds for $|H_p|$ less than any fixed constant, since otherwise \Lem{adjacent} implies we can choose $c$ 
 large enough such that $c|p|\log|p|$ will dominate all other terms.
 Next, observe that $\sum_{v\in p} \log|H_v| \leq \sum_{v\in p} \log(|H_v|+|p|) = |p|\log(|H_v|+|p|) = O(|p|\log|p|)$, as $p$ is a tall path.  
 Therefore, in the remainder of the proof  we can safely ignore the $\sum_{v\in p} \log|H_v|$ term as we can choose $c$ large enough to cover it.  
 That is we now instead prove the inequality,
 \[
  |H_p| \log |H_p|   \geq \frac{4}{3}\sum_{i=1}^k S_i \log S_i + \sum_{i=1}^l R_i \log R_i  -  c|p|\log |p|
 \]
 To do so we  break the analysis into two case based on the relative size of $|p|$ and $|H_p|$. 
 In the following, $\alpha>10$ is some yet to be determined constant.

\paragraph{Case 1, $|p|\geq |H_p|/\alpha$:\\}
First observe that by \Lem{adjacent}, $|H_p|+|p| \geq \sum_{i=1}^k S_i+ \sum_{i=1}^l R_i$.
By choosing $c$ large enough we therefore have,
\begin{align*}
 c|p|\log|p| &\geq 2(1+\alpha)^2 |p|\log |p| \geq 2(1+\alpha)|p| \log((1+\alpha)|p|)\\
             &\geq 2(|H_p|+|p|)\log(|H_p|+|p|) \geq 2\log(|H_p|+|p|) \pth{\sum_{i=1}^k S_i + \sum_{i=1}^l R_i}\\
             &\geq \frac{4}{3}\sum_{i=1}^k S_i \log S_i + \sum_{i=1}^l R_i \log R_i 
\end{align*}

In this case $\log(|H_p|+|p|)\leq \log(2|p|) \leq 1+\log(|p|)$ (where $\log$ is base $2$).
Therefore, assuming $|p|>1$, we have 
\[
(|H_p|+|p|)\log(|H_p|+|p|)\leq (1+c)|p|\log((1+c)|p|) %(|H_p|+|p|)(1+\log |p|) \leq 2(1+c)|p|\log |p|
\]
which when combined with the inequality from \Lem{adjacent} gives, 
\[
 4(1+c)|p|\log |p| \geq 2(|H_p|+|p|)\log(|H_p|+|p|)\geq \pth{|H_f| + \frac{4}{3}\sum_{i=1}^k S_i + \sum_{i=1}^l R_i}\log(|H_p|+|p|).
\]
This easily implies the desired inequality as $\log(H_p+p)$ is larger than any other $\log$ term in the inequality.

\paragraph{Case 2, $|p|< |H_p|/\alpha$:\\}

\Ben{Was not able to handle this case.}

%  First, lets assume we can show $\sum_{i=1}^k S_i + \sum_{i=1}^l R_i \leq \frac{3}{4}|H_p|$.  
%  In this case, using the last part of \Clm{sizes}, we can prove the desired inequality as follows.
%  
%  \begin{align*}
%   \frac{4}{3}\sum_{i=1}^k S_i \log S_i + \sum_{i=1}^l R_i \log R_i  &\leq  \frac{4}{3} \pth{\sum_{i=1}^k S_i \log S_i + \sum_{i=1}^l R_i \log R_i} \\
%   &\leq  \frac{4}{3} \log |H_p| \pth{\sum_{i=1}^k S_i + \sum_{i=1}^l R_i}  \leq  |H_p| \log |H_p|
%  \end{align*}
%  
%  Finally, to complete the proof we must prove our assumption that $\sum_{i=1}^k S_i + \sum_{i=1}^l R_i \leq \frac{3}{4}|H_p|$.
%  By \Lem{adjacent} and \Clm{sizes} we have, 
%  \[
%  \sum_{i=1}^k S_i + \sum_{i=1}^l R_i 
%  \leq |H_p|+|p|-|H_f| 
%  \leq (1+1/\alpha-1/3)|H_p| = (2/3+1/\alpha)|H_p|
%  \]
%  
%  Therefore $\sum_{i=1}^k S_i + \sum_{i=1}^l R_i \leq \frac{3}{4}|H_p|$ if $(2/3 + 1/\alpha)\leq 3/4$, which rearranged gives $\alpha \geq 12$
\end{proof}

\remove{
% \subsubsection{Paying it down the line}
% 
% \begin{lemma}
% \label{lem:excess}
%  Let $p$ be a tall path, and let $S_1, \dots S_k$ and $R_1, \dots, R_l$ be the sizes of the heaps at the roots of the short and tall children of $p$, respectively, in the shrub.  
%  Let $H_f$ be the size of the heap at the leaf of the path $p$, and let $X = \sum_{i=0}^k H_i \log H_i$.  
%  Then,  $|H_p| \log |H_p|   \geq  H_f\log H_f + \sum_{i=0}^k S_i \log S_i + \sum_{i=0}^l R_i \log R_i  -  c|p|\log |p|$, for some sufficiently large constant $c$
%  \end{lemma}
% \begin{proof}
% By \Lem{adjacent} we know that $|H_p|+|p| \geq  H_f + \sum_{i=0}^k S_i + \sum_{i=0}^l R_i$.  
% %Now as $p$ is a tall path we have $|p|\geq \sqrt{|H_p|}/100$, and so, $(|H_p|+|p|)\log(|H_p|+|p|)$
% 
% We  break the analysis into two case based on the relative size of $|p|$ and $|H_p|$.  In the following $c$ is some yet to be determined constant greater than $1$.
% 
% \paragraph{Case 1, $|p|\geq |H_p|/c$:}  
% In this case $\log(|H_p|+|p|)\leq \log(2|p|) \leq 1+\log(|p|)$ (where $\log$ is base $2$).
% Therefore, assuming $|p|>1$, we have 
% \[
% (|H_p|+|p|)\log(H_p+p)\leq (|H_p|+|p|)(1+\log |p|) \leq (2|p|)(2\log |p|)\leq 4|p|\log |p|
% \]
% which when combined with the inequality from \Lem{adjacent} gives, 
% \[
%  4|p|\log |p| \geq (|H_p|+|p|)\log(|H_p|+|p|)\geq \pth{H_f + \sum_{i=0}^k S_i + \sum_{i=0}^l R_i}\log(|H_p|+|p|).
% \]
% This easily implies the claim as $\log(H_p+p)$ is larger than any other $\log$ term in the inequality.
% 
% \paragraph{Case 2, $|p|< |H_p|/c$:}
% \end{proof}
}

Define the crust of the shrub to be the first (i.e. closest to the root) layer of short paths in the shrub.  Apply \Lem{shortCost} to each of the short paths in the crust and then 
say by \Lem{excess} and since the root heap goes to zero we are done.

\ignore{
Naturally, this geometric series behavior can be extended to more than two adjacent short paths, thus motivating 
the following definition.

\begin{definition}
 Let $p$ be a path in a path decomposition $P(T)$, with corresponding shrub $\pathTree$.
 Call a short path $q\in P(T)$ short reachable from $p$ if the path from $p$ to $q$ in $\pathTree$ 
 contains only vertices which correspond to short paths.  Call the subtree of $\pathTree$ consisting 
 of all short reachable paths from $p$ the \emph{short shrub} of $p$, and denote it $\mathcal{L}(p)$.
\end{definition}

In the following subsection we show how to exploit this geometric series behavior of short shrubs.

We say two paths in $P(T)$ are adjacent if their corresponding vertices are adjacent in the shrub.  Similarly, for 
two adjacent paths in $P(T)$ we can define a parent/child relationship based on this relationship in the shrub. 
}

\remove{
\Ben{Maybe move the remainder of this subsection to 5.3 since tall and short paths and not used until then}

\begin{definition}
 Let T be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.  
 For $p\in P(T)$, we say that $p$ is \emph{short} if $|p| < \sqrt{|H_{r_p}|}/100$, and we say it is \emph{tall} otherwise.
\end{definition}

\begin{observation}
\label{obs:decrease}
 Let $v$ be a vertex in $T$ and let $w$ be its parent.  Then $|H_w| \geq |H_v| -1$, as $L(T_v) \subseteq L(T_w)$ and the path from 
 $w$ to the root has one less vertex than the path from $v$ to the root.

 In particular, we have the following more general property.
 Let $v$ and $u$ be any two vertices in the same root to leaf path of $T$, such that $u$ is lower than $v$.  
 Then $|H_v| \leq |H_u| + d(u,v)$. 
\end{observation}

\begin{lemma}
\label{lem:pathBounds}
 Let T be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.
 Then, 
 \begin{compactenum}
  \item for any tall path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |p|)$.
  \item for any short path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |H_{r_p}|)$.
 \end{compactenum}
\end{lemma}
\begin{proof}
 By $\Obs{decrease}$ we know that for any vertex $v\in p$, $|H_v|\leq |H_{r_p}| + |p|$ (as $r_p$ lies below $v$ on $p$).
 As such, 
 \[
 \sum_{v\in p} \log |H_v| \leq \sum_{v\in p} \log(|H_{r_p}| + |p|)  = |p| \log (|H_{r_p}| + |p|).
 \]
 If $p$ is a tall path then $|p| \log (|H_{r_p}| + |p|) = O(|p| \log |p|)$, and if $p$ is short then 
 $|p| \log (|H_{r_p}| + |p|) = O(|p| \log |H_{r_p}| )$.
\end{proof}

The following lemma can be thought of as a generalization of $\Obs{decrease}$.

\begin{lemma}
\label{lem:adjacent}
 Let $T$ be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.
 Let $p$ be any path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the set of all adjacent child paths of $p$.
 For $1\leq i\leq k$ let $r_i = r_{q_i}$ and let $H_i = H_{r_i}$.
 Also, let $H_0$ denote the heap at the leaf of $p$.

 Then $\sum_{i=0}^k |H_i| \leq |H_p|+|p|$, where $H_p = H_{r_p}$.
\end{lemma}
\begin{proof} 
 First observe that for all $i$, $r_{i}$ lies above $r_p$ on some root to leaf path. 
 Let $H_i'$ be the subset of $H_i$ that lies below $r_p$.  Observe that $H_i'\subseteq H_p$, as $L(T_{r_i})\subseteq L(T_{r_p})$.

 Now for any $i\neq j$, $H_i \cap H_j = \emptyset$, as $r_{i}$ and $r_{j}$ are not on the same root to leaf path (i.e. $L(T_{r_i}) \cap L(T_{r_j}) = \emptyset$).
 In particular, $(H_i\setminus H_i') \cap (H_j \setminus H_j') = \emptyset$, and so $\sum_{i=0}^k |H_i\setminus H_i'| \leq |p|$ 
 as $H_i\setminus H_i'$ is precisely the subset of $H_i$ that lies on $p$.
 Moreover, for any $i\neq j$, $ H_i' \cap H_j' = \emptyset$, and so since $H_i'\subseteq H_p$ we have $\sum_{i=0}^k |H_i'|\leq |H_p|$.
 Putting these together gives
 \[
  \sum_{i=0}^k |H_i| = \sum_{i=0}^k |H_i'| + \sum_{i=0}^k |H_i\setminus H_i'| \leq |H_p|+|p|.
 \]
\end{proof}

% \Ben{There is a fundamental flaw in the following lemma and corollary.  
% As such, the remainder of this subsection will likely be deleted.}
% 
% Using the notation defined in the above lemma, we have the following crucial lemma, which we will use to 
% argue we can charge that cost of all paths in $P(T)$ to the cost of the tall paths.
% 
% \begin{lemma}
% Let $p$ be a tall path in $P(T)$ and let $\{q_1, \dots q_k\}$ be any set of short paths in $P(T)$ such that 
% $\sum_{i=1}^k |H_i| = O(H_p)$.  Then $\sum_{i=1}^k \sum_{v\in q_i} \log |H_v| = O(|p|\log |p|)$
% \end{lemma}
% \begin{proof}
% By \Lem{pathBounds}, for all $1\leq i\leq k$, 
%  $\sum_{v\in q_i} \log |H_v| = O(|q_i|\log |H_i|) = O(|H_i|\log |H_i|)$, as $q_i$ is a short path.
%  By assumption $\sum_{i=1}^k |H_i| = O(|H_p|)$, 
%  As $x\log x$ is a convex function, Jensen's inequality then implies 
%  $\sum_{i=1}^{k} |H_i|\log |H_i| = O(|H_p| \log |H_p|) = O(|p|\log |p|)$, since $p$ is a tall path.
% \end{proof}
% 
% 
% \begin{corollary}
% \label{cor:tallLight}
%  Let $p$ be a tall path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the set of all short paths adjacent to $p$.
%  Then $\sum_{i=1}^k \sum_{v\in q_i} \log |H_v| = O(|p|\log |p|)$.
% \end{corollary}
% \begin{proof}
%  Let $H_i$ be as defined in \Lem{adjacent}.  By \Lem{pathBounds}, for all $1\leq i\leq k$, 
%  $\sum_{v\in q_i} \log |H_v| = O(|q_i|\log |H_i|) = O(|H_i|\log |H_i|)$, as $q_i$ is a short path.
%  Now by \Lem{adjacent}, $\sum_{i=1}^k |H_i| \leq |H_p|+|p| = O(|p|)$, as $p$ is a tall path. 
%  Therefore, as $x\log x$ is a convex function, Jensen's inequality implies the claim.
%  
% \end{proof}

}

\subsection{The shrub of unsupported paths} \label{sec:short}

It suffices to focus on unsupported short paths. \Lem{short} can be reduced to the following.

\begin{lemma} \label{lem:support} $\sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| = O(\sum_{p \in \cL'} |H_p| \log |H_p|)$.
\end{lemma}

We now construct the shrub of unsupported short paths. Consider $p \in \cL \setminus \cL'$,
and traverse the chain of ancestors from $p$. Eventually, we must reach another short path $q$.
(If not, we have reached the root $r$ of $\pmax(T)$. Hence, $p$ is supported.)
Insert edge from $p$ to $q$, so $q$ is the parent of $p$ in $\cU$. This construction leads
to the shrub forest of $\cL \setminus \cL'$, where every unsupported short path is an internal node, and all
roots are supported short paths.

We list out some important claims about this shrub forest.
We focus on $\cU$, a connected component (shrub) in the shrub forest and denote the 
the root by $r(\cU)$.

\begin{claim} \label{clm:loss2} Consider $p \in \cL \setminus \cL'$ that is in $\cU$. Let $Q$
be the set of tall ancestors of $p$ that are descendants of $r(\cU)$. Then $\sum_{q \in Q} |q \cap H_p| \leq |H_p|/50$.
\end{claim}

\begin{proof} Suppose not, for contradictions sake. As we traverse the tall ancestors of $p$,
we encounter various vertices of $H_p$ in the ancestors $q$. Consider the first $|H_p|/100$
such vertices of $p$ encountered. At each such $v$, there are at least $|H_p|/100$
vertices \emph{not yet} encountered, so $|H_v| > |H_p|/100$. Each of these vertices $v$
is supported, implying that $p$ is a supported short path.
\end{proof}

\Lem{support} is a direct consequence of the following,
which is the really the main lemma.

%\begin{lemma} \label{lem:cu} Let $\cU$ denote a connected component (shrub) in the shrub forest of $\cL \setminus \cL'$ and let $r$
%be the root of $\cU$. Then $\sum_{p \in \cU} \sum_{v \in p} \log |H_v| = O(|H_r| \log |H_r|)$.
%\end{lemma}
%
%We proceed with the setup to prove \Lem{cu}.
%We focus our attention on some shrub $\cU$.
%We prove a technical lemma about $\cU$, from which \Lem{cu} follows directly.

\section{Ben's old stuff}

\begin{proof} (part (ii) of \Lem{cu-root}) Let level $\ell$ of $\cU$ denote all nodes at distance $\ell$ from the root $r$.
These set of these nodes is designated as $\cU_\ell$. We set $\sigma_\ell = \sum_{p \in \cU_\ell} |p|$
and $\sigma_{\leq \ell} = \sum_{k \leq \ell} \sigma_k$.
By \Lem{geometric}, for any $p \in \cU_\ell$, $|H_p| \leq (2/3)^\ell |H_r|$.

Since all nodes in $\cU$ represent short paths, $\forall p \in \cU$, $|p| < \sqrt{|H_p|}/100$.
So $\sigma_\ell < \sum_{p \in \cU_\ell} \sqrt{|H_p|}/100$.

\end{proof}

To prove this, we need to understand how $H_p$ is constructed. The path $p$ has some children
$q_1, q_2, \ldots, q_k$ in $\pmax(T)$. Furthermore, as a leaf path in $p$, some leaf
$\ell$ of $T$ in one of the ends of $p$. The heap $H_p$ is essentially the
union of $H_\ell$ and $H_{q_1}, H_{q_2}, \ldots, H_{q_k}$, with the vertices in $p$
removed. Note that some of the $q_i$'s are the corresponding $h(q')$ vertices,
for a child $q'$ in $\cU$.

>>>>

We present the first algorithm that can avoid the global sort and has a refined time complexity that
depends on the contour tree structure. Intuitively, if the tree is short and fat, we get significant improvements 
in running time. For a partition of the contour tree into a set of descending paths, $P$,
our algorithm runs in $O(\sum_{p\in P} |p|\log|p| + t\alpha(t)+N)$, which is 
always bounded by $O(t\log t + N)$ and is $O(t\alpha(t)+N)$ for balanced trees. 
Moreover, we prove a strong optimality property of our algorithm, showing that $\sum_{p\in P} |p|\log|p|$ complexity
is inherent to computing contour trees. Consider the family of instances where our algorithm
requires $C$ comparisons. Any algorithm computing contour trees on this family requires
$\Omega(C)$ comparisons in the worst case.

@@

% \paragraph{Organization.}  
% \Sec{basics} contains basic definition for contour trees.  In \Sec{approach} we give a high level overview and intuition for our approach 
% for computing contour trees.  \Sec{surgery} and \Sec{rain} describe how we break the intial input into nicely behaved (extemum dominant) pieces. 
% \Sec{extreme} shows that computing the contour tree for these nicely behaved pieces is roughly equivalent to computing the join tree. 
% \Sec{paint} contains our algorithm for computing join trees, which along with previous sections gives our algorithm for contour trees.  
% Finally, in \Sec{pathDecomp} and \Sec{lb} we relate the running time of our algorithm (providing upper and lower bounds) to a particular decomposition of the contour tree.

@@

The $(f(x)+\delta)$-contours in $f|_{B_\eps(x)}$ given by \Def{deg} might actually be the same
contour in $f$. The up-degree (as opposed to \emph{Morse} up-degree) is defined as the number of $(f(x)+\delta)$-contours
that intersect $B_\eps(x)$, a potentially smaller number. This up-degree
is exactly the up-degree of $x$ in $\reeb(\MM)$. (Analogously, for down-degree.)
When the Morse up-degree is $2$ but the up-degree is $1$, the topology of the level set changes but not by the number of connected components changing.
For example, when $d=3$ this is equivalent to the contour gaining a handle.
When $d=2$, this distinction is not necessary, since any point with $>1$
Morse degree will have $>1$ degree in $\reeb(\MM)$.

@@

\begin{figure}[h]\centering
    \includegraphics[width=.18\linewidth]{figs/tent2}%
    \hspace{.9in}
    \includegraphics[width=.18\linewidth]{figs/tent1}
    \caption{On left, downward rain spilling only (each shade of gray represents a piece created by each different spilling), producing a grid.  
    Note we are assuming raining was done in reverse sorted order of maxima.  
    On right, flipping the direction of rain spilling.}
    \label{fig:order}
\end{figure}

@@

We stress the importance of this step for our approach. Given the algorithm of~\cite{csa-cctad-00},
one may think that it suffices to design faster algorithms for join trees. But this
cannot give the sort of optimality we hope for. Again, consider \Fig{sorting}. Any algorithm
to construct the true join tree must construct the path of splits, which implies
sorting all of them. It is absolutely necessary to cut $\MM$ into pieces where the cost
of building the join tree can be related to that of building $\cC(\MM)$.

@@

\paragraph{The running time analysis:} Relating the running time to a path decomposition is the most technical part of the paper. 
All the non-heap operations can be easily bounded by $O(t\alpha(t) + N)$ (the $t\alpha(t)$ is from the union-find data structure
for colors). It is not hard to argue that at all times, any heap always contains a subset of a leaf
to root path. Unfortunately, this subset is \emph{not} contiguous, as the right part of \Fig{colors} shows. 
Specifically, in this figure the far left saddle (labeled $i$) is hit by blue paint.  However, there is another 
saddle on the far right (labeled $j$) which is not hit by blue paint.  Since this far right saddle is slightly 
higher than the far left one, it will merge into the component containing the blue mound (and also the yellow and red mounds)
before the far left one.  Hence, the vertices initially touched by blue are not contiguous in the contour tree.

Nonetheless, we can get a non-trivial (but non-optimal) bound. Let $d_v$ denote the distance to the root for vertex $v$ in the join tree.
The total cost (of the heap operations) is at most $\sum_v \log d_v$. This immediately proves a bound of $O(t\log D)$, where
$D$ is the maximum distance to the root, an improvement over previous work. But this bound is non-optimal.
For a balanced binary tree, this bound is $O(t\log\log t)$, whereas the cost of any path decomposition is $O(t)$.

@@

Consider a contour tree $T$ and the path decomposition $P(T)$ used to bound the running time. 
Denoting $\cost(P(T)) = \sum_{p \in P(T)} |p|\log |p|$, we construct a set of $\prod_{p \in P(T)} |p|!$
functions on a fixed domain such that each function has a distinct (labeled) contour tree. By a simple
entropy argument, any algebraic decision tree that correctly computes the contour tree on all instances
requires worst case $\Omega(\cost(P(T)))$ time. We prove that our algorithm makes $\Theta(\cost(P(T)))$ 
comparisons on all these instances. In general, any algorithm requires $\Omega(C)$ time
to solve the set of instances that our algorithm solves with $C$ comparisons.

@@

%\section{Acknowledgements}
\paragraph{Acknowledgements.}
We thank Hsien-Chih Chang, Jeff Erickson, and Yusu Wang for numerous useful discussions. This work is
supported by the Laboratory Directed Research and Development
(LDRD) program of Sandia National Laboratories. Sandia National
Laboratories is a multi-program laboratory managed and operated
by Sandia Corporation, a wholly owned subsidiary of Lockheed
Martin Corporation, for the U.S. Department of Energy’s National
Nuclear Security Administration under contract DE-AC04-94AL85000.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
%end of writup
%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@
